---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# susine

<!-- badges: start -->
<!-- badges: end -->

**susine** implements the Sum of Single Non-central Effects (SuSiNE) model, a
Bayesian variable-selection / fine-mapping method that extends
[SuSiE](https://github.com/stephenslab/susieR) by allowing nonzero, *signed*
prior means on effect sizes: `mu_0 = c * a`, where `a` is an annotation-derived
signed vector (e.g. from a sequence model) and `c` is a global scale. This lets
external functional information nudge the posterior toward biologically
plausible effects rather than only reweighting inclusion probabilities.

The package provides model fitting from:

- individual-level data — `susine()`,
- sufficient statistics — `susine_ss()`,
- summary statistics / RSS — `susine_rss()`,

together with the single-effect regression building blocks, empirical-Bayes
prior-update methods, and multi-start fit aggregation
(`aggregate_susine_fits()`). Each fit returns posterior inclusion probabilities
(PIPs), credible sets, and the evidence lower bound (ELBO).

## Installation

You can install the development version of susine from
[GitHub](https://github.com/Michael-G-Callahan/susine) with:

``` r
# install.packages("devtools")
devtools::install_github("Michael-G-Callahan/susine")
```

## Dependencies

`susine` is **pure R** (no compiled code). The core model depends only on
`Matrix` and `matrixStats`. In particular, **the core model does not require
`susieR`** — that dependency is only relevant to the paper's *baseline
comparisons*, which live in the companion `test-susine` repository. Those
comparisons pin the upstream package `stephenslab/susieR` (tag `0.15.53`,
commit `fbe5b7d`; the SuSiE-pathology exhibits use `0.15.51` / `34a4a52`) —
this is unmodified upstream SuSiE, not a fork.

## Example

A minimal fine-mapping fit on simulated data:

```{r example}
library(susine)

set.seed(1)
n <- 500
p <- 100

# Design matrix and a sparse truth with two causal variables
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
beta <- rep(0, p)
beta[c(10, 60)] <- c(1.5, -1.2)
y <- as.numeric(X %*% beta + rnorm(n))

# Fit SuSiNE with up to L = 5 single effects (default zero prior means
# recovers standard SuSiE-style behavior)
fit <- susine(L = 5, X = X, y = y)

# Posterior inclusion probabilities for the truly causal variables
fit$model_fit$PIPs[c(10, 60)]
```

To supply directional prior information, pass a signed annotation vector via
`mu_0` (and optionally let the global scale be fit with `auto_scale_mu_0 =
TRUE`). See `?susine`, `?susine_ss`, and `?susine_rss` for details.

## Reproducing the paper

This repository is the **model engine only**. The simulations, ensemble/scaling
analyses, sensitivity checks, and the real-data case study from the paper are
run through the companion **`test-susine`** harness — see its `REPRODUCIBILITY`
index for the paper-figure → method-trace → command map.
