---
title: "Development Testing Notebook"
author: "Michael Callahan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Development Testing Notebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(devtools)
load_all()
set.seed(1)
```

"The data-set contains a matrix of 574 individuals and 1,001 variables. These variables are real-world genotypes centered and scaled, and therefore retains the correlation structure of variables in the original genotype data. 3 out of the variables have non-zero effects. The response data is generated under a multivariate linear regression model. See Wang et al (2020) for more details."

```{r Get data}
data("SuSiE_N3_X") #This comes from the SuSiE pkg under 'data/'N3finemapping.rData'
# data(N3finemapping)
# 
# SuSiE_N3_X = N3finemapping$X
```

SuSiER simulated the response without showing the source code. Here is my take using their sparse sim code.

```{r Simualate y}
set.seed(1)

#Get n, p from X data
X = as.matrix(SuSiE_N3_X)[,1:100]
n = dim(X)[1]
p = dim(X)[2]

#Choose L
L = 6

sigma = 2 #make sure this doesn't dominate our dataset
# in the future, we will want to calc. the variance of Xb, then set
# the residual variance to be a small proportion of the systemic variance


#TODO: 
#Set the "true" params
beta = rep(0,p)
beta[1:5] = seq(-10,10, length.out = 5)

## SIMULATION: Turn into fcns
y = X %*% beta + rnorm(n, mean = 0, sd = sigma)

```

```{r Initialize arguments}

#Set prior params -- arbitrary for now
mu_0 = matrix(0,nrow = L, ncol = 1) 
sigma_0_2 = matrix(.1,nrow = L, ncol = 1) 
prior_inclusion_weights = matrix(1/p,nrow = 1,ncol = p)
# prior_inclusion_weights = matrix(c(1,rep(0,9)),nrow = 1,ncol = p)

```

```{r setup}
library(devtools)
# library(susieR)
library(matrixStats)
library(susieR)
load_all()
```
 
```{r}
undebug(susie)

#Fit susie to the raw data
susie_output <- susie(
    X,
    y,
    L
    )
susie_output$elbo
susie_output$pip

#Fit susine to the raw data
susine_output <- susine(
    L,
    X,
    y,
    prior_update_method = "var"
    )
susine_output$model_fit$PIPs
susine_output$model_fit$sigma_2
susine_output$model_fit$elbo

XtX = t(X)%*%X
Xty = t(X)%*%y
yty = sum(y^2)

#Fit susie_ss to the summary stats
susie_output <- susie_suff_stat(XtX, Xty, yty, n)
susie_output$elbo
susie_output$pip

#Fit susine_ss to the summary stats
susine_output <- susine_ss(
    L,
    XtX,
    Xty,
    yty,
    n=nrow(X),
    prior_update_method = "var"
    )

susine_output$model_fit$PIPs
susine_output$model_fit$sigma_2
susine_output$model_fit$elbo


#Using unstandardized z scores
sumstats <- univariate_regression(X, y)
z_scores <- sumstats$betahat / sumstats$sebetahat
R <- cor(X)
n=nrow(X)

susie_output  <- susie_rss(bhat = sumstats$betahat, 
                           shat = sumstats$sebetahat, 
                           n = n,
                           R = R, 
                           var_y = drop(var(y)),
                           L = 6,
                           estimate_residual_variance = TRUE)
susie_output$elbo
susie_output$pip

susine_output <- susine_rss(
    L,
    z_scores,
    R,
    n=n,
    s_hat = sumstats$sebetahat,
    var_y = drop(var(y)),
    prior_update_method = "var"
    )
susine_output$model_fit$PIPs
susine_output$model_fit$sigma_2
susine_output$model_fit$elbo

#Using standardized z scores -- ???
undebug(susie_rss)
susie_output  <- susie_rss(z = z_scores,
                           n = n,
                           R = R, 
                           L = 6,
                           estimate_residual_variance = TRUE)
susie_output$elbo
susie_output$sigma2

round(susie_output$pip,2)

undebug(susine_rss)
susine_output <- susine_rss(
    L,
    z_scores,
    R,
    n=n,
    prior_update_method = "var"
    )
susine_output$model_fit$PIPs
susine_output$model_fit$sigma_2
susine_output$model_fit$elbo

```
