---
title: "SuSiNE Tutorial: Priors, Scaling, and Compute Knobs"
date: 2025-10-13
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Overview

This tutorial walks through SuSiNE use cases with increasing amounts of prior information and optional autoscaling. It also shows "no‑prior" settings where you can trade extra compute for potentially better fits (annealing, multiple initializations).

```{r}
library(devtools)
load_all()

# utils::data("SuSiE_N3_X", package = "susine")
data("N3finemapping") # from susieR, under data/N3finemapping.rData
attach(N3finemapping)

library(susieR)
```

```{r}
# Simulation setup (one small region)
set.seed(1)
n = nrow(X)
p = ncol(X)
causal_idx = sample(seq_len(p), 5)
cat("Causal SNP indices:", paste(causal_idx, collapse=", "),"\n")

# Randomly generate effect sizes for these SNPs
b = rep(0, p)
b[causal_idx] = rnorm(n = 5, mean = 0, sd = 1)
cat("Causal effect sizes:", paste(round(b[causal_idx],2), collapse = ", "),"\n")

# Generate phenotype with target noise fraction
y_noise = 0.50
sigma_2 = var(X %*% b) * (y_noise / (1 - y_noise))
y = X %*% b + rnorm(n, mean = 0, sd = sqrt(sigma_2))

cat("Marginal effects at causals:", paste(round((as.vector(cov(X, y)) / diag(var(X)))[causal_idx],2), collapse = ", "),"\n")
invisible(cor(X[, causal_idx])) # show LD among causals
```

```{r}
# Create a vector of SNP annotations (prior means)
mu_0 = rnorm(n=p, mean = 0, sd = 0.1)
mu_0[causal_idx] = mu_0[causal_idx] + b[causal_idx] # make prior means close to true effects for causals

#boost the variance of the non-causals to match the causal
# mu_0[-causal_idx] = mu_0[-causal_idx]*(sd(mu_0[causal_idx])/sd(mu_0[-causal_idx]))

cat("Causal SNP annotations:", paste(round(mu_0[causal_idx],2), collapse = ", "),"\n")

#Show the variance of the annotations for causal SNPs
cat("Variance of causal SNP annotations:", round(var(mu_0[causal_idx]),4), "\n")

#Show the variance of the annotations for non-causal SNPs
cat("Variance of non-causal SNP annotations:", round(var(mu_0[-causal_idx]),4), "\n")


#Show the R^2 of the annotations for causal SNPs (annotation quality)
if (length(causal_idx) > 1 && sd(b[causal_idx]) > 0 && sd(mu_0[causal_idx]) > 0) {
  observed_r2 <- cor(mu_0[causal_idx], b[causal_idx])^2
} else {
  observed_r2 <- NA
}
cat("Annotation R^2 for causal SNPs:", round(observed_r2, 4), "\n")


# Create a vector of SNP annotations (prior variances)
sigma_0_2 = mu_0^2 # simple magnitude-only variance proxy
```

```{r}
# Metrics and helpers
if (!requireNamespace("PRROC", quietly = TRUE)) install.packages("PRROC")
library(PRROC)

truth <- integer(p); truth[causal_idx] <- 1L

pr_auc_vec <- function(scores, truth01) {
  pr <- PRROC::pr.curve(scores.class0 = scores[truth01 == 1], scores.class1 = scores[truth01 == 0], curve = FALSE)
  pr$auc.davis.goadrich
}

cross_entropy <- function(pip, truth01) {
  eps <- .Machine$double.eps
  pip <- pmin(pmax(pip, eps), 1 - eps)
  -mean(truth01 * log(pip) + (1 - truth01) * log(1 - pip))
}

h2_true <- var(as.vector(X %*% b)) / var(y)
cat(sprintf("True SNP Heritability (h_g^2): %.3f\n", h2_true))

h2_pred_from_fit <- function(fit) {
  s2 <- tail(fit$model_fit$sigma_2, 1)
  1 - s2 / var(y)
}

report_metrics <- function(fit, label) {
  pips <- fit$model_fit$PIPs
  cat(sprintf("[%s] ELBO: %.3f  Pred h2: %.3f  PR-AUC: %.3f  CE: %.3f\n",
              label,
              tail(fit$model_fit$elbo, 1),
              h2_pred_from_fit(fit),
              pr_auc_vec(pips, truth),
              cross_entropy(pips, truth)))
}

top_pips_table <- function(fit, k=10){
  pips = fit$model_fit$PIPs
  ord = order(pips, decreasing = TRUE)
  idx = head(ord, k)
  data.frame(SNP_Index = idx,
             PIP = round(pips[idx],2),
             True_Selection = as.integer(idx %in% causal_idx))
}
```

---
# Part A: No User Priors
Start with no prior information and explore options from naive to EB. Then add compute-heavy knobs (annealing, multiple inits).

```{r}
### a.i. SuSiE — Naive mu, naive sigma (no EB)
cat("--- Running: a.i (naive mu/sigma) ---\n")
# undebug(susine)
fit_a_i <- susine(L=5, X, y, prior_update_method = "none")
report_metrics(fit_a_i, "a.i naive mu/sigma")
test <- susie(X, y, L=5, estimate_prior_variance = FALSE)

# Largest PIP difference vs susieR
pip_diffs  = test$pip - fit_a_i$model_fit$PIPs
max_diff_idx = which(pip_diffs == max(pip_diffs))
paste("SNP with largest PIP difference:", max_diff_idx)
paste("SuSiE value:", round(test$pip[max_diff_idx],4))
paste("SuSiNE value:",  round(fit_a_i$model_fit$PIPs[max_diff_idx],4))
paste("Difference:", round(pip_diffs[max_diff_idx],4))
```

```{r}
### a.ii. SuSiE — EB sigma, naive mu
cat("\n--- Running: a.ii (EB sigma) ---\n")
fit_a_ii <- susine(L=5, X, y, prior_update_method = "var")
report_metrics(fit_a_ii, "a.ii EB sigma, naive mu")

test <- susie(X, y, L=5, estimate_prior_variance = TRUE)
pip_diffs  = test$pip - fit_a_ii$model_fit$PIPs
max_diff_idx = which(abs(pip_diffs) == max(abs(pip_diffs)))
paste("SNP with largest PIP difference:", max_diff_idx)
paste("SuSiE value:", round(test$pip[max_diff_idx],4))
paste("SuSiNE value:",  round(fit_a_ii$model_fit$PIPs[max_diff_idx],4))
paste("Difference:", round(pip_diffs[max_diff_idx],4))
max(abs(pip_diffs))
```

```{r}
### a.iii. SuSiNE — EB mu, naive sigma
cat("\n--- Running: a.iii (EB mu) ---\n")
fit_a_iii <- susine(L=5, X, y, prior_update_method = "mean")
report_metrics(fit_a_iii, "a.iii EB mu, naive sigma")
```

```{r}
### a.iv. SuSiNE — EB mu and EB sigma
cat("\n--- Running: a.iv (EB mu & EB sigma) ---\n")
fit_a_iv <- susine(L=5, X, y, prior_update_method = "both")
report_metrics(fit_a_iv, "a.iv EB mu & EB sigma")
```

```{r}
### a.v. No priors + annealing
# Slows updates early to reduce greedy moves; potentially better ELBO
fit_a_v <- susine(L=5, X, y,
                  prior_update_method = "var",
                  anneal_start_T = 100,
                  anneal_schedule_type = "geometric",
                  anneal_target = "both",
                  anneal_burn_in = 30)
report_metrics(fit_a_v, "a.v naive + annealing")
```

```{r}
### a.vi. No priors + multiple random inits (ensemble and best-ELBO)

L_susine <- 5

## Calibrate init_random_sd analytically so the aggregate start explains ~5% of Var(y)
target_var_frac <- 0.1           # target aggregate signal variance as fraction of Var(y)
X_for_init <- initialize_X(X)     # reuse SuSiNE's standardization to access attr(X,"d")
sum_d <- sum(attr(X_for_init, "d"))
total_target_var <- target_var_frac * var(y)
per_effect_target_var <- total_target_var / L_susine
sd_cal <- sqrt(per_effect_target_var * (n - 1) / sum_d)

cat(sprintf("Calibrated per-effect init_random_sd: %.4g (aggregate target %.1f%% of Var(y))\n",
            sd_cal, 100 * target_var_frac))

set.seed(1)
multi_init_runs <- lapply(1:10, function(seed){
  susine(L=L_susine, X, y,
         prior_update_method = "none",
         init_random = TRUE,
         init_random_sd = sd_cal,
         init_seed = seed)
})

# Aggregate PIPs across runs (equal and ELBO-weighted)
agg_equal <- aggregate_susine_fits(multi_init_runs, weighting = "equal")
agg_elbo  <- aggregate_susine_fits(multi_init_runs, weighting = "elbo", temperature = 1)

# Best single run by ELBO for comparison
elbos <- vapply(multi_init_runs, function(f) tail(f$model_fit$elbo,1), numeric(1))
best_idx <- which.max(elbos)
fit_a_vi_best <- multi_init_runs[[best_idx]]

report_metrics(fit_a_vi_best, sprintf("a.vi best of 10 inits (seed %d)", best_idx))
report_metrics(agg_equal, "a.vi ensemble (equal)")
report_metrics(agg_elbo,  "a.vi ensemble (ELBO-weighted)")
```

---
# Part B: User Priors (Unscaled)
Provide informative priors without autoscaling.

```{r}
### b.i. SuSiE — functional sigma (mu=0)
cat("\n--- Running: b.i (functional sigma; fixed) ---\n")
functional_sigma <- mu_0^2 / var(y) # proportion of Var(y)
fit_b_i <- susine(L=5, X, y, mu_0 = 0,
                  sigma_0_2 = functional_sigma,
                  prior_update_method = "none",
                  auto_scale_mu_0 = FALSE,
                  auto_scale_sigma_0_2 = FALSE)
report_metrics(fit_b_i, "b.i mu=0, functional sigma (no autoscale)")
```

```{r}
### b.ii. SuSiNE — functional mu, EB sigma
cat("\n--- Running: b.ii (functional mu; EB sigma) ---\n")
fit_b_ii <- susine(L=5, X, y,
                   mu_0 = mu_0,
                   prior_update_method = "var",
                   auto_scale_mu_0 = FALSE,
                   auto_scale_sigma_0_2 = FALSE)
report_metrics(fit_b_ii, "b.ii functional mu (no autoscale), EB sigma")
```

```{r}
### b.iii. SuSiNE — functional mu and functional sigma (fixed)
cat("\n--- Running: b.iii (functional mu & sigma; fixed) ---\n")

# Demonstration: fix sigma_0_2 via ground truth calibration (scalar)
sigma0_sq_hat <- mean((b[causal_idx] - mu_0[causal_idx])^2)
sigma0_prop_hat <- as.numeric(sigma0_sq_hat / var(y))

fit_b_iii <- susine(L=5, X, y,
                    mu_0 = mu_0,
                    sigma_0_2 = rep(sigma0_prop_hat, p),
                    prior_update_method = "none",
                    auto_scale_mu_0 = FALSE,
                    auto_scale_sigma_0_2 = FALSE)
report_metrics(fit_b_iii, "b.iii functional mu & sigma (no autoscale)")
```

---
# Part C: User Priors with Autoscaling
Autoscaling helps when priors are directionally/magnitude‑informative but mis‑scaled overall.

```{r}
### c.i. Functional sigma + autoscale variance
fit_c_i <- susine(L=5, X, y, mu_0 = 0,
                  sigma_0_2 = functional_sigma,
                  prior_update_method = "none",
                  auto_scale_mu_0 = FALSE,
                  auto_scale_sigma_0_2 = TRUE)
report_metrics(fit_c_i, "c.i mu=0, functional sigma + autoscale sigma")
```

```{r}
### c.ii. Functional mu + autoscale mean (variance EB)
fit_c_ii <- susine(L=5, X, y,
                   mu_0 = mu_0,
                   prior_update_method = "var",
                   auto_scale_mu_0 = TRUE,
                   auto_scale_sigma_0_2 = FALSE)
report_metrics(fit_c_ii, "c.ii functional mu + autoscale mu; EB sigma")

#print the fitted mu scale
fit_c_ii$priors$mu_0_scale_factor
```

```{r}
### c.iii. Functional mu & sigma + autoscale both
fit_c_iii <- susine(L=5, X, y,
                    mu_0 = mu_0,
                    sigma_0_2 = functional_sigma,
                    prior_update_method = "none",
                    auto_scale_mu_0 = TRUE,
                    auto_scale_sigma_0_2 = TRUE)
report_metrics(fit_c_iii, "c.iii functional mu & sigma + autoscale both")
```

```{r}
# Optional: show top PIPs for a few cases
cat("\nTop PIPs: a.i (naive)\n"); top_pips_table(fit_a_i)
cat("\nTop PIPs: b.i (functional sigma)\n"); top_pips_table(fit_b_i)
cat("\nTop PIPs: c.i (functional sigma + autoscale)\n"); top_pips_table(fit_c_i)
```


```{r}
# --- given ---
# ρ-level credible set by cumulative PIP
get_credible_set <- function(alpha, cutoff = 0.95) {
  if (!is.numeric(alpha)) stop("alpha must be numeric (PIPs).")
  if (!is.finite(cutoff) || cutoff <= 0 || cutoff >= 1) stop("cutoff must be in (0,1).")
  a <- alpha
  a[!is.finite(a)] <- 0
  if (sum(a) <= 0) return(integer(0))  # nothing to select

  # order by decreasing PIP; break ties by original index
  o  <- order(-a, seq_along(a))
  cs <- cumsum(a[o])
  k  <- which(cs >= cutoff)[1]
  if (is.na(k)) integer(0) else o[seq_len(k)]
}

# --- NEW: per-effect metrics using your purity/coverage definitions ---
cs_metrics_one_effect <- function(alpha_vec, X, causal_idx, cutoff = 0.95) {
  cs <- get_credible_set(alpha_vec, cutoff = cutoff)
  sz <- length(cs)

  # purity = min |r| among pairs in CS (size-1 -> 1.0; empty -> NA)
  purity <- if (sz >= 2) {
    C <- cor(X[, cs, drop = FALSE], use = "pairwise.complete.obs")
    min(abs(C[upper.tri(C)]))
  } else if (sz == 1) {
    1.0
  } else {
    NA_real_
  }

  # coverage: does this CS contain at least one true causal SNP?
  coverage <- as.integer(sz > 0 && any(cs %in% causal_idx))

  list(indices = cs, size = sz, purity = purity, coverage = coverage)
}

# --- NEW: wrapper across all L effects in fit_c_ii ---
summarize_model_cs <- function(fit, X, causal_idx, cutoff = 0.95) {
  stopifnot(is.matrix(X))
  L <- fit$settings$L
  A <- fit$effect_fits$alpha   # expected: L x p

  eff <- vector("list", L)
  for (l in seq_len(L)) {
    eff[[l]] <- cs_metrics_one_effect(A[l, ], X, causal_idx, cutoff = cutoff)
  }

  effects_df <- data.frame(
    effect   = seq_len(L),
    size     = sapply(eff, `[[`, "size"),
    purity   = sapply(eff, `[[`, "purity"),
    coverage = sapply(eff, `[[`, "coverage"),
    indices  = I(lapply(eff, `[[`, "indices"))
  )

  # union across effects for power
  cs_union <- sort(unique(unlist(effects_df$indices)))
  power <- if (length(causal_idx) > 0) {
    length(intersect(cs_union, causal_idx)) / length(causal_idx)
  } else NA_real_

  model_summary <- data.frame(
    L             = L,
    mean_size     = mean(effects_df$size, na.rm = TRUE),
    mean_purity   = mean(effects_df$purity, na.rm = TRUE),
    mean_coverage = mean(effects_df$coverage, na.rm = TRUE),
    power         = power
  )

  list(effects = effects_df, model = model_summary, cs_union_indices = cs_union)
}

# --- example usage (replaces your single-effect snippet) ---
res <- summarize_model_cs(
  fit = fit_c_ii,
  X = X,                       # n x p
  causal_idx = causal_idx,     # integer indices of true causal SNPs
  cutoff = 0.95
)

res$effects      # per-effect table (indices is a list-column)
res$model        # model-level aggregates (means + power)
res$cs_union_indices

```

## Oracle ELBO Diagnostic

```{r}
# Force the single-effect updates to stay on the true causal SNPs and record the ELBO path
oracle_idx <- rep(causal_idx, length.out = fit_a_i$settings$L)

run_oracle_ibss <- function(fit_ref, X, y, idx_vec, n_iter = 10, sigma2_init = NULL) {
  L_ref <- fit_ref$settings$L
  idx_vec <- rep(idx_vec, length.out = L_ref)
  stopifnot(all(idx_vec >= 1L), all(idx_vec <= ncol(X)))
  prior_eps <- sqrt(.Machine$double.eps)
  diag_mode <- TRUE

  # Standardize inputs using the same utilities as the main fit
  y_cent <- y - mean(y)
  X_std <- initialize_X(X)
  n <- nrow(X_std)
  p <- ncol(X_std)
  L <- L_ref

  # Extract per-effect prior parameters so we can reuse the same scales
  mu_mat <- fit_ref$priors$mu_0
  sigma_mat <- fit_ref$priors$sigma_0_2/100
  prior_mat <- fit_ref$priors$prior_inclusion_weights

  sigma2_trace <- numeric(n_iter + 1)
  sigma2_seed <- tail(fit_ref$model_fit$sigma_2[is.finite(fit_ref$model_fit$sigma_2)], 1)
  if (length(sigma2_seed) == 0 || !is.finite(sigma2_seed)) {
    sigma2_seed <- var(y_cent)
  }
  sigma2_trace[1] <- if (!is.null(sigma2_init)) sigma2_init else sigma2_seed
  sigma2_lb <- if (!is.null(fit_ref$settings$residual_variance_lowerbound)) {
    fit_ref$settings$residual_variance_lowerbound
  } else {
    var(y_cent) / 1e4
  }

  effect_b_hat <- matrix(0, nrow = L, ncol = p)
  effect_b2_hat <- matrix(0, nrow = L, ncol = p)
  KL_vec <- rep(0, L)
  elbo_trace <- numeric(n_iter + 1)
  elbo_trace[1] <- get_objective(X_std, y_cent, effect_b_hat, effect_b2_hat, sigma2_trace[1], KL_vec)

  for (iter in seq_len(n_iter)) {
    sigma2_current <- sigma2_trace[iter]
    if (diag_mode && !is.finite(sigma2_current)) {
      cat(sprintf("diag: sigma2_current NA at iter %d\n", iter))
    }
    full_residuals <- y_cent - compute_Xb(X_std, colSums(effect_b_hat))
    if (diag_mode && any(!is.finite(full_residuals))) {
      cat(sprintf("diag: full_residuals has NA at iter %d (first NA index: %s)\n",
                  iter,
                  paste(which(!is.finite(full_residuals))[1], collapse = ",")))
    }

    for (l in seq_len(L)) {
      partial_residuals <- full_residuals + compute_Xb(X_std, effect_b_hat[l, ])
      j <- idx_vec[l]
      if (diag_mode && any(!is.finite(partial_residuals))) {
        cat(sprintf("diag: partial_residuals NA at iter %d, effect %d\n", iter, l))
      }

      l_idx <- min(l, nrow(mu_mat))
      mu_vec <- mu_mat[l_idx, ]
      sigma_vec <- sigma_mat[l_idx, ]
      prior_vec <- prior_mat[l_idx, ]

      xty_full <- compute_Xty(X_std, partial_residuals)
      d_j <- attr(X_std, "d")[j]
      xty_j <- xty_full[j]
      if (diag_mode && !is.finite(xty_j)) {
        cat(sprintf("diag: xty_j NA at iter %d, effect %d, SNP %d\n", iter, l, j))
      }
      if (diag_mode && !is.finite(d_j)) {
        cat(sprintf("diag: d_j NA at iter %d, effect %d, SNP %d\n", iter, l, j))
      }

      denom <- sigma2_current + sigma_vec[j] * d_j
      if (diag_mode && (!is.finite(denom) || denom == 0)) {
        cat(sprintf("diag: denom issue (%.3g) at iter %d, effect %d, SNP %d (sigma2=%.3g, sigma_vec=%.3g, d_j=%.3g)\n",
                    denom, iter, l, j, sigma2_current, sigma_vec[j], d_j))
      }
      if (diag_mode && (!is.finite(sigma_vec[j]) || sigma_vec[j] < 0)) {
        cat(sprintf("diag: sigma_vec problem at iter %d, effect %d, SNP %d (sigma_vec=%.3g)\n",
                    iter, l, j, sigma_vec[j]))
      }
      mu1_j <- (sigma2_current * mu_vec[j] + sigma_vec[j] * xty_j) / denom
      sigma1_j <- (sigma2_current * sigma_vec[j]) / denom
      if (diag_mode && (!is.finite(mu1_j) || !is.finite(sigma1_j))) {
        cat(sprintf("diag: posterior moment NA at iter %d, effect %d, SNP %d (mu1=%.3g, sigma1=%.3g)\n",
                    iter, l, j, mu1_j, sigma1_j))
      }

      effect_b_hat[l, ] <- 0
      effect_b_hat[l, j] <- mu1_j
      effect_b2_hat[l, ] <- 0
      effect_b2_hat[l, j] <- sigma1_j + mu1_j^2

      log_BF_vec <- BF(X_std, partial_residuals, sigma2_current, sigma_vec, mu_vec)
      log_prior_j <- log(prior_vec[j] + prior_eps)
      lpo_j <- log_BF_vec[j] + log_prior_j
      if (diag_mode && (!is.finite(log_BF_vec[j]) || !is.finite(lpo_j))) {
        cat(sprintf("diag: log terms NA at iter %d, effect %d, SNP %d (logBF=%.3g, log_prior=%.3g)\n",
                    iter, l, j, log_BF_vec[j], log_prior_j))
      }

      ser_erss <- SER_ERSS(X_std, partial_residuals, effect_b_hat[l, ], effect_b2_hat[l, ])
      base_loglik <- sum(dnorm(partial_residuals, 0, sqrt(sigma2_current), log = TRUE))
      KL_vec[l] <- - (lpo_j + base_loglik) -
        ((n / 2) * log(2 * pi * sigma2_current) + (1 / (2 * sigma2_current)) * ser_erss)
      if (diag_mode && !is.finite(KL_vec[l])) {
        cat(sprintf("diag: KL NA at iter %d, effect %d (base_loglik=%.3g, lpo_j=%.3g, ser_erss=%.3g)\n",
                    iter, l, base_loglik, lpo_j, ser_erss))
      }

      full_residuals <- partial_residuals - compute_Xb(X_std, effect_b_hat[l, ])
    }

    elbo_trace[iter + 1] <- get_objective(X_std, y_cent, effect_b_hat, effect_b2_hat, sigma2_current, KL_vec)
    sigma2_candidate <- ERSS(X_std, y_cent, effect_b_hat, effect_b2_hat) / n
    if (diag_mode && !is.finite(elbo_trace[iter + 1])) {
      cat(sprintf("diag: ELBO NA at iter %d\n", iter))
    }
    if (diag_mode && !is.finite(sigma2_candidate)) {
      cat(sprintf("diag: sigma2_candidate NA at iter %d\n", iter))
    }
    sigma2_trace[iter + 1] <- max(sigma2_candidate, sigma2_lb)
    if (diag_mode && !is.finite(sigma2_trace[iter + 1])) {
      cat(sprintf("diag: sigma2_trace NA after update at iter %d (candidate=%.3g, lower=%.3g)\n",
                  iter, sigma2_candidate, sigma2_lb))
    }
  }

  list(
    idx = idx_vec,
    elbo = elbo_trace,
    sigma2 = sigma2_trace,
    b_hat = effect_b_hat,
    b_2_hat = effect_b2_hat,
    KL = KL_vec,
    X_std = X_std,
    y_cent = y_cent
  )
}

oracle_run <- run_oracle_ibss(
  fit_ref = fit_a_i,
  X = X,
  y = y,
  idx_vec = oracle_idx,
  n_iter = 10,
  sigma2_init = sigma_2
)

oracle_elbo <- tail(oracle_run$elbo, 1)
oracle_elbo_sigma_updated <- get_objective(
  oracle_run$X_std,
  oracle_run$y_cent,
  oracle_run$b_hat,
  oracle_run$b_2_hat,
  tail(oracle_run$sigma2, 1),
  oracle_run$KL
)
fit_elbo <- tail(fit_a_i$model_fit$elbo[is.finite(fit_a_i$model_fit$elbo)], 1)

cat(sprintf("Oracle ELBO (forced causal support, sigma2 used in sweep): %.3f\n", oracle_elbo))
cat(sprintf("Oracle ELBO (after updating sigma2): %.3f\n", oracle_elbo_sigma_updated))
cat(sprintf("fit_a_i ELBO: %.3f\n", fit_elbo))

oracle_path <- data.frame(
  iteration = seq_along(oracle_run$elbo) - 1,
  sigma2 = oracle_run$sigma2,
  elbo = oracle_run$elbo
)

oracle_erss <- ERSS(oracle_run$X_std, oracle_run$y_cent, oracle_run$b_hat, oracle_run$b_2_hat)

list(
  elbo_path = oracle_path,
  oracle_erss = oracle_erss,
  oracle_sigma2 = tail(oracle_run$sigma2, 1)
)
```

## Truth-Initialized Trajectory

```{r}
truth_idx <- causal_idx

L_truth <- fit_a_i$settings$L
truth_idx_rep <- rep(truth_idx, length.out = L_truth)

X_std_for_init <- initialize_X(X)
scale_vec <- attr(X_std_for_init, "original:scale")
scale_vec[scale_vec == 0] <- 1

alpha_init <- matrix(0, nrow = L_truth, ncol = p)
b_hat_init <- matrix(0, nrow = L_truth, ncol = p)
for (l in seq_len(L_truth)) {
  j <- truth_idx_rep[l]
  alpha_init[l, j] <- 1
  b_hat_init[l, j] <- b[j] * scale_vec[j]
}

y_centered_truth <- y - mean(y)
resid_truth_std <- y_centered_truth - compute_Xb(X_std_for_init, colSums(b_hat_init))
sigma2_init_truth <- var(resid_truth_std)
cat(sprintf("sigma2 (nominal simulation): %.6f  |  sigma2 (standardized truth residuals): %.6f\n",
            sigma_2, sigma2_init_truth))

fit_truth <- susine(
  L = L_truth,
  X = X,
  y = y,
  prior_update_method = fit_a_i$settings$prior_update_method,
  anneal_start_T = fit_a_i$settings$anneal_start_T,
  anneal_schedule_type = fit_a_i$settings$anneal_schedule_type,
  anneal_target = fit_a_i$settings$anneal_target,
  anneal_burn_in = fit_a_i$settings$anneal_burn_in,
  prior_update_warmup = fit_a_i$settings$prior_update_warmup,
  init_alpha = alpha_init,
  init_sigma2 = sigma2_init_truth,
  record_history = TRUE
)

pip_history <- fit_truth$model_fit$pip_history
elbo_vec <- fit_truth$model_fit$elbo
sigma_vec <- fit_truth$model_fit$sigma_2

metrics_truth <- data.frame(
  iteration = seq_along(elbo_vec) - 1,
  elbo = elbo_vec,
  sigma2 = sigma_vec,
  auprc = vapply(pip_history, function(pip) pr_auc_vec(pip, truth), numeric(1)),
  cross_entropy = vapply(pip_history, function(pip) cross_entropy(pip, truth), numeric(1))
)

metrics_truth
```

```{r}
library(ggplot2)

ggplot(metrics_truth, aes(x = elbo, y = auprc)) +
  geom_path(color = "grey70") +
  geom_point(size = 2.2) +
  geom_text(aes(label = iteration), vjust = -0.8, size = 3.5) +
  labs(
    title = "AUPRC trajectory from truth-initialized IBSS",
    x = "ELBO",
    y = "PR-AUC (PIPs vs truth)"
  )
```

```{r}
ggplot(metrics_truth, aes(x = elbo, y = cross_entropy)) +
  geom_path(color = "grey70") +
  geom_point(size = 2.2) +
  geom_text(aes(label = iteration), vjust = -0.8, size = 3.5) +
  labs(
    title = "Cross-Entropy trajectory from truth-initialized IBSS",
    x = "ELBO",
    y = "Cross-Entropy (PIPs vs truth)"
  )
```
